What's in a Name? Linear Temporal Logic Literally Represents Time Lines

### 2023-06-26 paper/main.tex
typos/minor changes
more typos and reviewer comments

\subsection{State-based \Buchi automata (BA)}
\begin{definition}[$\omega$-word]
    An $\omega$-word or infinite run of $\Sigma$, is an infinite string $s = (s_0, s_1, s_2, \dots)$ where each $s_i\in \Sigma$
\end{definition}
\begin{definition}[\Buchi automaton]
    A \Buchi automaton is a $5$-tuple, $G = (Q, \Sigma, \delta, s, F)$ consisting of
    \begin{enumerate}
        \item a finite set of states $Q$
        \item a finite alphabet of input symbols called the alphabet $\Sigma$
        \item a partial function $\delta : Q\times \Sigma \to Q$, called the transition function
        \item an initial or start state called $s\in Q$
        \item a set of accepting states $F \subseteq Q$
    \end{enumerate}
    $G$ accepts an infinite run iff at least one of its infinitely visited states is in $F$
\end{definition}

\subsection{$\omega$-regular expression}
\begin{definition}[Regular expression]\label{def:regex}
    For word $a \in \Sigma$, regular expression is defined by the following grammar:
    \begin{align*}
        A ::= \emptyset \mid \epsilon \mid a \mid AA \mid A + A \mid A^*
    \end{align*}
\end{definition}
\begin{definition}[Semantics of regular expression]
    Let $L(A)$ denote the language accepted by regular expression $A$. Then $L(A)$ is inductively defined as
    \begin{align*}
        L(\emptyset) & = \emptyset \\
        L(\epsilon) & = \{\epsilon\} \tag{$\epsilon$ denotes the empty string}\\
        L(a) & = \{a\} \\
        L(A_1A_2) & = \{s_1s_2 \mid s_1 \in L(A_1) \text{ and } s_2 \in L(A_2)\} \\
        L(A_1 + A_2) & = L(A_1) \cup L(A_2) \\
        L^{(0)}(A) & = \{\epsilon\} \\
        L^{(i + 1)}(A) & = \{s_1s_2 \mid s_1 \in L(A) \text{ and } s_2 \in L^{(i)}(A)\} \\\
        L(A^*) & = \bigcup_{i \ge 0} L^{(i)}(A)
    \end{align*}
\end{definition}

@@

\begin{definition}[$\omega$-regular expression]\label{def:omega-regex}
    Regular expression concerns only finite-length strings. However, since \ltl\ formulas reason about events that happen over an infinite-length timeline, we need to model it using infinite regular expressions (i.e. $\omega$-regular expression). $\omega$-regular expression is defined by the following grammar:
    \begin{align*}
        B ::= A^{\omega} \mid AB \mid B + B
    \end{align*}
\end{definition}
\begin{definition}[Semantics of $\omega$-regular expression]\label{def:omega-semantics}
    Let $\Sigma^{\omega}$ denote the set of infinite-length string over fixed alphabet $\Sigma$. Let $L_{\omega}(B)$ denote the $\omega$-language accepted by $\omega$-regular expression $B$. Then $L_{\omega}(B)$ is inductively defined as
    \begin{align*}
        L_{\omega}(A^{\omega}) & = \{s_1s_2s_3\cdots \mid s_i \in L(A) \text{ and } i \ge 1\} \tag{$\epsilon
         \not\in L(A)$} \\
        L_{\omega}(AB) & = \{s_1s_2 \mid s_1 \in L(A) \text{ and } s_2 \in L_{\omega}(B)\} \\
        L_{\omega}(B_1 + B_2) &= L_{\omega}(B_1) \cup L_{\omega}(B_2)
    \end{align*}
\end{definition}

\begin{remark}
  \Cref{def:regex,def:omega-regex} are the standard definitions of regular expressions and $\omega$-regular expressions, respectively. We include them here for completeness.
\end{remark}

@@

\subsection{BA to $\omega$-regex} \label{aut2regex}
We translate \Buchi automata to $\omega$-regular expressions by first finding the regular expressions for paths from the start state to some final state (say $r_\mathrm{sf}$), and for those for paths for looping from the final state back to itself (say $r_\mathrm{ff}$), and finally combining those to form the $\omega$-expression for valid runs ($\bigplus_{f\in F} r_\mathrm{sf}r_\mathrm{ff}^\omega$). The regular expressions for finite paths are in turn found by iteratively deleting interior nodes in digraph of the automaton.

@@

\begin{algorithm}[h!]
    \label{alg:ba2wregex}
    \caption{ba2wregex}
    \begin{algorithmic}
        \Require $G$, a \Buchi automaton
        \Ensure The $\omega$-regular expression recognized by $G$
        \State \Return $\bigcup_{f \in F}\left(\text{nfa2regex\_firstvisit}(G, s, f)\text{\^{}nfa2regex}(G, f, f) \right) $
    \end{algorithmic}
\end{algorithm}

@@

\section{Theoretical Analyses} \label{sec:analysis}
\subsection{Correctness of Regex Translation}
\begin{lemma}
    % reduce_nfa does its input to output behaviour
    For any NFA $G$ and any state $v$ in $G$ that is neither a start state or a final state, \textbf{reduce\_nfa}$(G, v)$ preserves the regular language accepted by $G$
\end{lemma}
\begin{proof}
    Let $G'$ be the graph of $G$ post reduction by the application of \textbf{reduce\_nfa}$(G, v)$. We show that the trace of every path accepted by $G$ is also accepted by $G'$. Suppose a path accepted by $G$ does not pass through $v$, clearly this is true. Else, suppose it passes through $v$, since $v$ is an interior node, $v$ cannot be the first or last in the path. Thus, for every pass through $v$, let $u\neq v$ be the last node passed before entering $v$, and likewise $w$ be the first node after exiting $v$. We show that the regular language of sub-traces from $u$ to $w$ in $G$ (shown in \cref{fig:uvw-dfa}) is identical to that in $G'$.
    \begin{figure}[h!]
        \centering
        \begin{tikzpicture}
	\node[draw, thick, shape=ellipse, style={minimum width=1.5cm, minimum height=1cm}] (u) {$u$};
	\node[draw, thick, shape=ellipse, style={minimum width=1.5cm, minimum height=1cm}] (v) [right=of u] {$v$};
	\node[draw, thick, shape=ellipse, style={minimum width=1.5cm, minimum height=1cm}] (w) [right=of v] {$w$};
	\draw[thick] (u) edge["$r_{uv}$",-latex] (v) (v) edge["$r_{vv}$",-latex,loop above] (v) (v) edge["$r_{vw}$",-latex] (w);
        \end{tikzpicture}
 %       \includegraphics[scale=0.75]{img/uvw_dfa.pdf}
        \caption{A path from $u$ to $w$ in $G$}
        \label{fig:uvw-dfa}
    \end{figure}
    Suppose there is no loop at $v$, the path from $u$ to $w$ must simply be $u\xrightarrow{r_{uv}}v\xrightarrow{r_{vw}}w$, with the trace $r=r_{uv}r_{vw}$. If there is a loop edge $r_{vv}$, this edge can be traveled any number of times before exiting $v$, thus the regular expression is $r=r_{uv}r_{vv}^*r_{vw}$. This edge $u\xrightarrow{r}w$ is added in $G'$. Thus, for every path accepted by $G$, for every subpath entering and exiting $v$ from some $u$ to $w$, there exists a corresponding subpath from $u$ to $w$ in $G'$, and thus a corresponding path accepted by $G'$.
\end{proof}

>>>>

\subsection{State-based \Buchi automata (BA)}
\begin{definition}[$\omega$-word]
    An $\omega$-word or infinite run of $\Sigma$, is an infinite string $s = (s_0, s_1, s_2, \dots)$ where each $s_i\in \Sigma$
\end{definition}
\begin{definition}[\Buchi automaton]
    A \Buchi automaton is a $5$-tuple, $G = (Q, \Sigma, \delta, s, F)$ consisting of
    \begin{enumerate}
        \item a finite set of states $Q$
        \item a finite alphabet of input symbols called the alphabet $\Sigma$
        \item a relation $\delta \subseteq (Q\times \Sigma) \times Q$, called the transition relation
        \item an initial or start state called $s\in Q$
        \item a set of accepting states $F \subseteq Q$
    \end{enumerate}
    $G$ accepts an infinite run iff at least one of its infinitely visited states is in $F$
\end{definition}

\subsection{$\omega$-regular expression}
\begin{definition}[Regular expression]\label{def:regex}
    For symbol $a \in \Sigma$, regular expression is defined by the following grammar:
    \begin{align*}
        A ::= \emptyset \mid \epsilon \mid a \mid AA \mid A + A \mid A^*
    \end{align*}
\end{definition}
\begin{definition}[Semantics of regular expression]
    Let $L(A)$ denote the set of propositional formulas accepted by regular expression $A$. Then $L(A)$ is inductively defined as
    \begin{align*}
        L(\emptyset) & = \emptyset \\
        L(\epsilon) & = \{\epsilon\} \tag{$\epsilon$ denotes the empty string}\\
        L(a) & = \{a\} \\
        L(A_1A_2) & = \{s_1s_2 \mid s_1 \in L(A_1) \text{ and } s_2 \in L(A_2)\} \\
        L(A_1 + A_2) & = L(A_1) \cup L(A_2) \\
        L^{(0)}(A) & = \{\epsilon\} \\
        L^{(i + 1)}(A) & = \{s_1s_2 \mid s_1 \in L(A) \text{ and } s_2 \in L^{(i)}(A)\} \\\
        L(A^*) & = \bigcup_{i \ge 0} L^{(i)}(A)
    \end{align*}
\end{definition}

@@

\begin{definition}[$\omega$-regular expression]\label{def:omega-regex}
    Regular expression concerns only finite-length strings. However, since \ltl\ formulas reason about events that happen over an infinite-length timeline, we need to model it using infinite regular expressions (i.e. $\omega$-regular expression). $\omega$-regular expression is defined by the following grammar:
    \begin{align*}
        B ::= A^{\omega} \mid AB \mid B + B
    \end{align*}
\end{definition}
\begin{definition}[Semantics of $\omega$-regular expression]\label{def:omega-semantics}
    Let $\Sigma^{\omega}$ denote the set of infinite-length strings over fixed alphabet $\Sigma$. Let $L_{\omega}(B)$ denote the $\omega$-language accepted by $\omega$-regular expression $B$. Then $L_{\omega}(B)$ is inductively defined as
    \begin{align*}
        L_{\omega}(A^{\omega}) & = \{s_1s_2s_3\cdots \mid s_i \in L(A) \text{ and } i \ge 1\} \tag{$\epsilon
         \not\in L(A)$} \\
        L_{\omega}(AB) & = \{s_1s_2 \mid s_1 \in L(A) \text{ and } s_2 \in L_{\omega}(B)\} \\
        L_{\omega}(B_1 + B_2) &= L_{\omega}(B_1) \cup L_{\omega}(B_2)
    \end{align*}
\end{definition}

\begin{remark}
  \Cref{def:regex,def:omega-regex} are the standard definitions of regular expressions and $\omega$-regular expressions, respectively.  We include them here for completeness. For the remainder of the paper, we denote arbitrary regular expressions using $A$ and $\omega$-regular expressions by $B$. 
\end{remark}

@@

\subsection{BA to $\omega$-regex} \label{aut2regex}
We translate \Buchi automata to $\omega$-regular expressions by first finding the regular expressions for paths from the start state to some final state (say $r_\mathrm{sf}$), and for those for paths looping from the final state back to itself (say $r_\mathrm{ff}$), and finally combining those to form the $\omega$-expression for valid runs ($\bigplus_{f\in F} r_\mathrm{sf}r_\mathrm{ff}^\omega$). The regular expressions for finite paths are in turn found by iteratively deleting interior nodes in the digraph of the automaton.

@@

\begin{algorithm}[h!]
    \label{alg:ba2wregex}
    \caption{ba2wregex}
    \begin{algorithmic}
        \Require $G$, a \Buchi automaton
        \Ensure The $\omega$-regular expression recognized by $G$
        \State \Return $\bigcup_{f \in F}\left(\text{nfa2regex\_firstvisit}(G, s, f)\text{nfa2regex}(G, f, f) \right) $
    \end{algorithmic}
\end{algorithm}

@@

\section{Theoretical Analyses} \label{sec:analysis}
In the following section, we prove the correctness of our translation pipeline. The currectness of the translation from LTL formulas to \Buchi automata is provided by SPOT~\cite{Dur22}. Below, we prove the correctness of (i) the translation from \Buchi automata to $\omega$-regular expressions, (ii) the ($\omega$-)regular expression rewrite rules we apply, and lastly, we outline the correctness of our timeline visualizations. 
\subsection{Correctness of Regex Translation}
\begin{lemma}
    % reduce_nfa does its input to output behaviour
    For any NFA $G$ and any state $v$ in $G$ that is neither a start state or a final state, \textbf{reduce\_nfa}$(G, v)$ preserves the regular language accepted by $G$
\end{lemma}
\begin{proof}
    Let $G'$ be the graph of $G$ post reduction by the application of \textbf{reduce\_nfa}$(G, v)$. We show that the trace of every path accepted by $G$ is also accepted by $G'$. Suppose a path accepted by $G$ does not pass through $v$, clearly this is true. Else, suppose it passes through $v$, since $v$ is an interior node, $v$ cannot be the first or last in the path. Thus, for every pass through $v$, let $u\neq v$ be the last node passed before entering $v$, and likewise $w$ be the first node after exiting $v$. We show that the regular language of sub-traces from $u$ to $w$ in $G$ (shown in \cref{fig:uvw-dfa}) is identical to that in $G'$.
    \begin{figure}[h!]
        \centering
        \begin{tikzpicture}
	\node[draw, thick, shape=ellipse, style={minimum width=1.5cm, minimum height=1cm}] (u) {$u$};
	\node[draw, thick, shape=ellipse, style={minimum width=1.5cm, minimum height=1cm}] (v) [right=of u] {$v$};
	\node[draw, thick, shape=ellipse, style={minimum width=1.5cm, minimum height=1cm}] (w) [right=of v] {$w$};
	\draw[thick] (u) edge["$r_{uv}$",-latex] (v) (v) edge["$r_{vv}$",-latex,loop above] (v) (v) edge["$r_{vw}$",-latex] (w);
        \end{tikzpicture}
 %       \includegraphics[scale=0.75]{img/uvw_dfa.pdf}
        \caption{A path from $u$ to $w$ in $G$}
        \label{fig:uvw-dfa}
    \end{figure}
    Suppose there is no loop at $v$, the path from $u$ to $w$ must simply be $u\xrightarrow{r_{uv}}v\xrightarrow{r_{vw}}w$, with the trace $r=r_{uv}r_{vw}$. If there is a loop edge $r_{vv}$, this edge can be traveled any number of times before exiting $v$, thus the regular expression is $r=r_{uv}r_{vv}^*r_{vw}$. This edge $u\xrightarrow{r}w$ is added in $G'$. Thus, for every path accepted by $G$, for every subpath entering and exiting $v$ from some $u$ to $w$, there exists a corresponding subpath from $u$ to $w$ in $G'$, and thus a corresponding path accepted by $G'$.
\end{proof}
